---
title: 'Week1: Review of Basics'
author: "Brent H."
date: "2025-01-27"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # FALSE hides the code but shows the output (if any)
knitr::opts_chunk$set(message = FALSE) # FALSE suppresses messages in the output document.
knitr::opts_chunk$set(warning = FALSE) # FALSE suppresses warnings in the output document.

options(scipen=999)  # Set to display small numbers in decimal form
set.seed(1234) # Optional, for reproducibility

# Check and load required libraries
# tidyverse: A collection of packages for data manipulation and visualization
if(!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)

# ggplot2: A system for creating graphics
if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)

# haven: for reading in stata (.dta) files
if(!require(haven)) install.packages("haven")
library(haven) 

# modelsummary: Summarizes and presents regression models and data frames in tables
if(!require(modelsummary)) install.packages("modelsummary")
library(modelsummary)

```

#### Let's refresh our memory of the RStudio Environment

```{r}
getwd() # Show the working directory

# Use setwd() if you need to change the file path or point-and-click in the file pane to the right
```

## R Basics Review
Let's start by creating a vector of whole random numbers called `a`.
```{r}
a <- sample(1:100, size = 10, replace = T)
```

Here are some basic functions for measures of central tendency.

Mode:
```{r}
getmode <- function(v) {   # getmode is a function we create because the "mode" function does something else
      uniqv <- unique(v)       # create a vector `uniqv` that has only unique values of `v`
      uniqv[which.max(tabulate(match(v, uniqv)))]  # follows the steps from by-hand method above 
}

getmode(a)
```

Median:
```{r}
median(a)
```

Mean:
```{r}
mean(a)
```

Here are some basic functions for measures of variability.

Range:
```{r}
range(a)
```

Inter-quartile range
```{r}
IQR(a)
```

Variance and standard deviation
```{r}
var(a)

sd(a)
```

## Let's load the data

Download wage1 dataset from Github into your working directory then import wage1.dta using the `read_dta` function from the haven package. Name the object `wage_org`.

```{r}
wage_org <- read_dta("DATASETS/wage1.dta")
```

Info about dataset: http://fmwww.bc.edu/ec-p/data/wooldridge/datasets.list.html
Codebook: http://fmwww.bc.edu/ec-p/data/wooldridge/wage1.des

```{r}
# Copy wage_org to a working dataset called wage1.
wage1 <- wage_org
```

### Basic Data Overview

#### Exercise

Inspect data in R viewer using `View` function.

```{r}
View(wage1)
```

*Question 1*: What do the columns and rows represent?

Response*:* Columns represent variables or characteristics of people, who are represented as rows; each row represents one person (unit) observed.

*Question 2a*: Using code, how many rows and columns are there?

```{r}
dim(wage1)
```

*Question 2b*: How big is the sample size? (n = ?)

Response: n = 526

*Question 3a*: Find participant 35 in the View(wage1), how many years of education did they report?

Response: 12 years of education

*Question 3b*: Instead of manually looking up the row and column, what code works to answer this?

```{r}
wage1[35,"educ"]
```

#### Exercise

Use the `str` function to understand the structure of the dataset.

```{r}
str(wage1)
```

*Question 4*: Are there any factor variables?

Response: No, there aren't any factor variables.

Question 5: Determine the number of missing values in each variable.

```{r}
colSums(is.na(wage1))
```

#### Exercise

-   Use the `head` function to display the first few rows of the dataset. This gives a glimpse of the actual data values.

```{r}
head(wage1)
```

### Data Recoding & Wrangling

#### Exercise

-   Create a new factor variable called `gender` based on the `female` variable. This variable should have two levels: "Male" and "Female", where `female == 1` is "Female".

```{r}
wage1 <- wage1 %>%
  mutate(gender = factor(ifelse(female == 1, "Female", "Male"), 
                         levels = c("Male", "Female")))
```

#### Exercise

Recode the `educ` variable into a new categorical variable `educ_level` as a factor variable with the following levels:
    - "Less than High School" (less than 12 years)
    - "High School Graduate" (12 years)
    - "Some College" (13 to 15 years)
    - "College Graduate and above" (16 or more years)

```{r}
wage1 <- wage1 %>%
  mutate(educ_level = factor(case_when(
    educ < 12 ~ "Less than High School",
    educ == 12 ~ "High School Graduate",
    educ > 12 & educ < 16 ~ "Some College",
    educ >= 16 ~ "College Graduate and above"
  ), levels = c("Less than High School", "High School Graduate", "Some College", "College Graduate and above")))

```

#### Exercise

Create a new factor variable `region` based on `northcen`, `south`, and `west`. The regions should be labeled as 'North', 'South', 'West'. If a participant does not fall into any of these categories, label them as `Other`.

```{r}
wage1 <- wage1 %>%
  mutate(region = case_when(
    northcen == 1 ~ "North Central",
    south == 1 ~ "South",
    west == 1 ~ "West",
    TRUE ~ "Other"  # default case
  )) %>%
  mutate(region = factor(region))
```

*Follow-up*: Did you check if there are any participants who were categorized in more than one region? If there are, how will you handle these cases?

```{r}
wage1 %>%
  filter(northcen + south + west > 1) %>%
  nrow()
```

#### Exercise

Select a subset of columns for further analysis (`wage`, `educ_level`, `exper`, `tenure`, `gender`, `regio`) from the `wage1` dataset and name the new data frame `wage2`.

```{r}
wage2 <- wage1 %>%
  select(wage, educ_level, exper, tenure, gender, region)  
```

### Descriptive Statistics

#### Exercise

Calculate the mean, median, and standard deviation for `wage`,`exper`, and `tenure`.

```{r}
mean(wage2$wage, na.rm = TRUE)
median(wage2$wage, na.rm = TRUE)
sd(wage2$wage, na.rm = TRUE)

mean(wage2$exper, na.rm = TRUE)
median(wage2$exper, na.rm = TRUE)
sd(wage2$exper, na.rm = TRUE)

mean(wage2$tenure, na.rm = TRUE)
median(wage2$tenure, na.rm = TRUE)
sd(wage2$tenure, na.rm = TRUE)
```

#### Exercise

Calculate frequency tables with counts and percentages for the categorical variables `educ_level`, `gender`, and `region`. Combine the results for each variable into a single table.

Education level:
```{r}
educ_freq <- table(wage2$educ_level)
educ_percent <- prop.table(educ_freq) * 100
educ_table <- cbind(Count = educ_freq, Percentage = round(educ_percent,2))
educ_table
```

Gender:
```{r}
gender_freq <- table(wage2$gender)
gender_percent <- prop.table(gender_freq) * 100
gender_table <- cbind(Count = gender_freq, Percentage = round(gender_percent, 2))
gender_table
```

Region:

```{r}
region_freq <- table(wage2$region)
region_percent <- prop.table(region_freq) * 100
region_table <- cbind(Count = region_freq, Percentage = round(region_percent, 2))
region_table
```

*Question 6*: What is the most common education level in the dataset?

Response: High school graduate with 37.64% of sample.

*Question 7*: What is the average number of years of experience (`exper`) among the participants?

```{r}
mean(wage2$exper) 
# Response: 17.017 years of experience
```

*Question 8*: Which education level has the highest average hourly earnings?

```{r}
wage2 %>%
  group_by(educ_level) %>%
  summarise(
    Mean = mean(wage, na.rm = TRUE)
  )
# Response: College graduates and above with $8.95/hourly earnings
```

*Question 9*: What is the standard deviation of the education level with the second highest average hourly earnings?

```{r}
wage2 %>%
  group_by(educ_level) %>%
  summarise(
    Mean = mean(wage, na.rm = TRUE),
    SD = sd(wage,na.rm = TRUE)
  )
# Response: Some College Mean = 6.03 and SD = 3.32
```

*Question 10*: Based on the descriptive statistics obtained, what kind of statistical analysis would you perform to make a more formal inference about the relationship between education level and wages? Why would you choose this method?

Response: ANOVA. The DV is interval ratio and IV is composed of more than two educational categories.

### Data Visualizations

#### Exercise

Fill in the blanks in the below code for a box-and-whiskers plot.

```{r}
ggplot(wage2, aes(x = educ_level, y = wage)) +
  geom_boxplot() +
  labs(title = "Figure 1. Average Hourly Earnings by Education Level",
       x = "Education Level",
       y = "Hourly Earnings",
       caption = "Notes: n=526; Wooldridge Data Set") +
  theme_minimal()
```

#### Exercise
Fill in the blanks in the below code for a histogram of average wages grouped by education level.

Calculate the mean wage for each education level.
```{r}
mean_wages <- wage2 %>%
  group_by(educ_level) %>%
  summarise(Mean = mean(wage, na.rm = TRUE))
```

Create the histogram with density plots and mean lines.
```{r}
ggplot(wage2, aes(x = wage, fill = educ_level)) +
  geom_density(alpha = 0.7) +
  geom_vline(data = mean_wages, aes(xintercept = Mean, color = educ_level), linetype = "dashed") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Figure 2. Average Hourly Earnings by Education Level",
       x = "Hourly Earnings",
       y = "Density",
       caption = "Notes: n=526; Wooldridge Data Set") +
  theme_minimal()
```

Nice work :) End of Week 1 Lab
