---
title: "Week 5 - Categorical and Nonlinear Predictors"
author: "SOC542: Spring 2026"
date: "2026-02-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen = 999)

library(tidyverse)
library(haven)
library(modelsummary)
library(rstanarm)
library(tidybayes)

seed <- 1234
set.seed(seed)
```

# Data Preparation: GSS 2022

We will use the General Social Survey (GSS) to investigate how career and demographic variables affect personal income.

```{r data-load-clean}
# Load and initial cleaning
gss_raw <- read_dta("DATASETS/GSS/GSS2022.dta")

gss <- gss_raw %>%
  select(conrinc, hrs1, prestg10, degree, sex, race, age, partyid, wtssnrps) %>%
  zap_label() %>%
  mutate(
    # Clean continuous variables
    income = as.numeric(conrinc),
    prestige = as.numeric(prestg10),
    age = ifelse(age < 0, NA, age),
    workhrs = case_when(
      hrs1 == 89 ~ NA, # Remove outliers
      hrs1 < 0  ~ NA,
      TRUE      ~ as.numeric(hrs1)
    ),
    
    # Nonlinear & Scaling transformations
    log_inc = log(income),
    workexp = age - 18,
    workexp_2 = (age - 18)^2,
    
    # Recode categorical variables as factors
    # Note: R uses the first level of a factor as the reference category in regressions.
    # You can change this using relevel(factor_variable, ref = "Desired_Reference").
    degree = factor(degree, levels = 0:4, labels = c("Less_HS", "HS", "Assoc", "Bach", "Grad")),
    sex = factor(sex, levels = 1:2, labels = c("Male", "Female")),
    female = relevel(sex, ref = "Female"),
    race = factor(race, levels = 1:3, labels = c("White", "Black", "Other")),
    party = case_when(
      partyid %in% 0:2 ~ "Democrat",
      partyid %in% 4:6 ~ "Republican",
      TRUE             ~ "Independent"
    ) %>% factor(levels = c("Democrat", "Independent", "Republican")),
    
  ) %>%
  drop_na(income, age, workhrs, prestige, sex, race)

datasummary_balance(~ sex, data = gss, title = "Descriptive Statistics")
```

# Part 1: Categorical Predictors (Dummies)

When we include a factor in `lm()`, R automatically creates dummy variables and drops the first level as the reference category.

```{r categorical-models}
# Model with Male, Black, and Democrat as reference groups
m1 <- lm(income ~ age + workhrs + prestige + sex + race + party, data = gss)
m2 <- lm(income ~ age + workhrs + prestige + female + race + party, data = gss)

modelsummary(list("Ref: Male" = m1, "Ref: Female" = m2), 
             stars = TRUE, 
             coef_map = c("(Intercept)" = "(Intercept)",
                          "sexFemale" = "Sex: Female",
                          "femaleMale" = "Sex: Male",
                          "age" = "age",
                          "workhrs" = "workhrs",
                          "prestige" = "prestige",
                          "raceBlack" = "raceBlack",
                          "raceOther" = "raceOther",
                          "partyIndependent" = "partyIndependent",
                          "partyRepublican" = "partyRepublican"),
             gof_omit = "F|RMSE|Log|IC")
```

# Part 2: Nonlinearity & Multiplicative Relationships
## Log-transformation of DV

While logarithms are often used to fix skewed distributions, their more important **theoretical** use is for modeling **multiplicative relationships**. 

In variables like income or population, we often expect effects to be proportional rather than additive. For example, a "1-unit increase" in education shouldn't just add a fixed dollar amount to everyone's salary; instead, it might increase a salary by a certain *percentage*. (Also think of the population example!) Logarithms allow us to model this multiplicative nature using linear regression.

By modeling $\ln(Y) = \beta_0 + \beta_1 X_1 + \epsilon$, we are implying that:
$$Y = e^{\beta_0 + \beta_1 X_1 + \epsilon}$$
$$Y = e^{\beta_0} \cdot e^{\beta_1 X_1} \cdot e^{\epsilon}$$

In this exponentiated form, the relationship is clearly multiplicative. Each 1-unit increase in $X_1$ multiplies the expected value of $Y$ by a factor of $e^{\beta_1}$ (See https://library.virginia.edu/data/articles/interpreting-log-transformations-in-a-linear-model for more information).

## Adding a squared term of IDV

We often use **squared terms** (a **second-order polynomial**) to model relationships whose slopes are not constant. In sociology, this is common for "life-cycle" effects, where an outcome (like income) increases with age but at a diminishing rate, eventually leveling off or declining.


```{r nonlinear-models}
# Comparing Linear vs Log vs Polynomial
ols_linear <- lm(income ~ workexp + workhrs + sex, data = gss)
ols_log    <- lm(log_inc ~ workexp + workhrs + sex, data = gss)
ols_poly   <- lm(income ~ workexp + workexp_2 + workhrs + sex, data = gss)

modelsummary(list("Linear" = ols_linear, "Log" = ols_log, "Polynomial" = ols_poly),
             stars = TRUE, 
             coef_map = c("(Intercept)" = "(Intercept)",
                          "workexp" = "workexp",
                          "workexp_2" = "workexp_2",
                          "workhrs" = "workhrs",
                          "sexFemale" = "sexFemale"),
             gof_omit = "F|RMSE|Log|IC", 
             title = "Transformation Comparison")
```

### Visualizing the Squared Effect
```{r plot-nonlinear}
# Extract coefficients for the curve: y = b0 + b1x + b2x^2
b <- coef(ols_poly)
curve_func <- function(x) b["(Intercept)"] + b["workexp"]*x + b["workexp_2"]*x^2
vertex = b["workexp"] / (-2*b["workexp_2"])

ggplot(gss, aes(x = workexp, y = income)) +
  stat_function(fun = curve_func, color = "red", linewidth = 1) +
  geom_vline(aes(xintercept=vertex)) +
  labs(title = "Diminishing Returns of Work Experience", x = "Years of Experience", y = "Income") +
  theme_minimal()
```

# Part 3: Survey Weights & Bayesian Estimation

Finally, we account for sampling probability using weights and compare our results with a Bayesian approach.

```{r final-comparison}
# OLS with Weights
ols_weighted <- lm(income ~ workexp + workexp_2 + prestige + sex + race, 
                   data = gss, weights = wtssnrps)

# Bayesian with Weights
bayes_weighted <- stan_glm(income ~ workexp + workexp_2 + prestige + sex + race, 
                           data = gss, weights = wtssnrps, 
                           seed = seed, refresh = 0, chains = 1)

modelsummary(list("OLS (Weighted)" = ols_weighted, "Bayes (Weighted)" = bayes_weighted),
             stars = TRUE,
             statistic = "conf.int",
             fmt=2,
             gof_omit = "F|RMSE|Log|IC|alg|pss")
```

# Exercise
Specify a model that predicts `log_inc` using `degree` and other controls.
1. Which degree level is the reference category by default?
2. Use `relevel()` to make "Grad" the reference category. 
3. Interpret the coefficient for "Less_HS" in the new model.

```{r exercise}
gss = gss %>%
  mutate(degree_relevel = relevel(BLANK, ref = BLANK))

your_model = lm(log_inc ~ BLANK + BLANK, data=gss)

summary(your_model)


```