---
title: "Week2_Linear Regression"
author: "YJ Chae"
date: "2026-02-02"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # FALSE hides the code but shows the output (if any)
knitr::opts_chunk$set(message = FALSE) # FALSE suppresses messages in the output document.
knitr::opts_chunk$set(warning = FALSE) # FALSE suppresses warnings in the output document.

options(scipen=999)  # Set to display small numbers in decimal form
set.seed(1234) # Optional, for reproducibility

# Load required libraries
library(haven) # For reading .dta files
library(tidyverse) # For data manipulation
library(modelsummary) # For really nice descriptive statistics
library(stargazer) # For nice looking tables
library(ggplot2) # For visualizations

```

Most Stata datasets contain informative variables labels that you can view using R functions. We'll learn how to view them and store them into an R object.

To take advantage of these labels, download a dataset in Stata format if available and load it using `read_dta` function in the `haven` package.

## Loading the Dataset

```{r}
crime <- read_dta("DATASETS/DC_2000_crime_data.dta")
```

## Exploratory Data Analysis

Viewing the first few rows of the dataset

```{r}
head(crime)

```

*Question:* Provide a general characterization of the dataset. What kind of information does it contain?

*Answer:* 

Let's look at the data using the data viewer.

```{r}
view(crime) # Note that tidyverse provides a handy `view` function as an alternative to `View` in base R.

```

*Question:* Did you notice the small text under the variable names in the data viewer? What does the label say for the variable 'per_nohs'?

*Answer:* 

## Extracting and learning more about the available variables

Extract variable names into a new object called `var_names`

```{r}
var_names <- names(crime)
print(var_names)
```

#### Exercise: 

Initialize an empty vector to store variable names and labels.

```{r}
var_labels <- c()

```

Let's take a look at the label for the `per_tanf` variable.

```{r}
attr(crime[["per_tanf"]], "label")

```

Instead of loading the label one by one, here is a `for` loop that goes through each variable to extract and store its label.

```{r}
for(var in var_names) {
  var_labels[which(var_names == var)] <- attr(crime[[var]], "label")
}

```

Create a data frame with variable names and labels

```{r}
variables_df <- data.frame(
  Variable = var_names,
  Label = var_labels,
  stringsAsFactors = FALSE  # to keep the text as characters
)

```

View the definition of the available variables in the dataset

```{r}
view(variables_df)

```

Generating some quick descriptive statistics of the `crime` dataset

```{r}
datasummary_skim(crime)
```

*Question:* Identify the variable in the dataset that has approximately 4% missing values.

*Answer:*

#### Exercise:

Create a histogram for visualizing the distribution of violent crime rate in 2001 by filling in the **BLANK** with the variable name of interest

```{r}
ggplot(BLANK, aes(x = BLANK)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Violent Crime Rate in 2001",
       x = "Violent Crime Rate (per 1,000)",
       y = "Frequency")
```

Create a histogram of the violent crime rate in 2001 with bin increments of 25

```{r}
ggplot(BLANK, aes(x = BLANK)) +
  geom_histogram(binwidth = BLANK, fill = "blue", color = "black") +
  labs(title = "Histogram of Violent Crime Rate in 2001",
       x = "Violent Crime Rate (per 1,000)",
       y = "Frequency")
```

*Question:* How does changing the binwidth to 25 affect the visualization?
Is the plot improved?

*Answer:* 


## Identify a Bivariate Relationship

**Exploring the relationship between `poverty_rate` and `vcr2001` in a scatter plot**

Fill in the **BLANK**. Also find out the appropriate `geom_...` function for scatter plot.

```{r}
crime %>%
  ggplot(aes(x = BLANK, y = BLANK)) +
  geom_...(color="black") +
  ggtitle("Scatter Plot of Violent Crime Rate vs. % Poverty") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty", 
       caption="Notes: Washington D.C. census tracts, 2000") +
  theme_light()
```

*Question:* What is your impression of the relationship between % poverty and violent crime rate?
Is it positive, negative, or non-linear?

*Answer:* 

**Visual Representation of Total Sum of Squares (TSS)**

```{r}
mean_vcr <- mean(crime$vcr2001, na.rm = TRUE)

crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  # Add horizontal line representing the overall mean
  geom_hline(yintercept = mean_vcr, linetype = "dashed", color = "blue") +
  # Add vertical lines representing distance from each point to the mean
  geom_segment(aes(xend = poverty_rate, yend = mean_vcr), 
               color = "red", alpha = 0.5) +
  ggtitle("Scatter plot of violent crime rate and % poverty") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty", 
       caption = "Notes: Washington D.C. census tracts, 2000. Red lines represent distances contributing to TSS") +
  theme_minimal()
```

*Question:* In the above plot, the red lines represent the distance of each observation from the overall mean of violent crime rate.
To calculate the Total Sum of Squares (TSS) an important concept for regression analysis, what mathematical operation should you perform on these distances?
What does TSS tell us about the data's variability?

*Answer:* 

#### Exercise: 

Complete the scatter plot with line of best fit for `poverty_rate` and `vcr2000`.
Fill in the **BLANK** with appropriate code

```{r}
crime %>%
  ggplot(aes(x = BLANK, y = BLANK))+
  geom_point(color="black") +
  geom_smooth(method="lm", color="red", fill=NA) + # Note the method "lm" for geom_smooth!
  ggtitle("Scatter plot of violent crime rate and % poverty") +
  labs(y="VCR (per 1,000)", x="% Poverty", caption="Notes: Washington D.C. census tracts, 2000") 
```

#### Exploring the relationship between `homeownership_rate` and `vcr2001`

#### Exercise: 

Create a Scatter Plot of `vcr2001` and `homeownership_rate` with line of best fit and averages

```{r}
crime %>%
  ggplot(aes(x = BLANK, y = BLANK))+
  geom_point(color="black") +
  geom_smooth(method="BLANK", color="red", fill=NA)+
  ggtitle("Scatter Plot of Violent Crime Rate vs. % Homeownership with Averages") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Homeownership", caption="Notes: Washington D.C. census tracts, 2000") 
```

*Question:* After creating the plot, describe the relationship between % homeownership and violent crime rate.

*Answer:* 

## Simple Linear Regression

Fitting a linear model: Using the `poverty_rate` to predict `vcr2001`

```{r}
lm_model <- lm(BLANK ~ BLANK, data = crime)

```

Displaying the summary of the linear model

#### Exercise: 

Debug the code below

```{r}
sum(lm_model)
```

*Question:* Based on the above output, what is the estimated change in violent crime rate for each additional unit increase in poverty rate (poverty_rate)?

*Answer:* 

*Question:* What is the significance of the intercept in this model?
What does it represent in the context of violent crime rate?

*Answer:* 

#### *Exercise:* 

Create another linear model exploring the relationship between `homeownership_rate` and `vcr2001`

```{r}
lm_model2 <- lm(BLANK ~ BLANK, data = crime)
summary(lm_model2)
```

Use `modelsummary` for a more visually appealing presentation of models (especially in reports or presentations)

```{r}
modelsummary(lm_model2,
             stars = T,
             statistic = "std.error", # experiment with statistic = "conf.int"
             fmt=3, # Control the number of decimal places to show using this argument.
             notes = "Notes: Add your note here.") 

```

*Question:* Report your findings

*Answer:*

## Predicted Values and Residuals

We will use the linear model `lm_model(vcr2001 ~ poverty_rate)`. Based on our modeling, let's predict crime rates using `poverty_rate` as the independent variable.

Generating Predicted Values

```{r}
crime$predicted_vcr <- predict(lm_model)

```

Histograms of Actual and Predicted Violent Crime Rates Objective:

Create two overlaid histograms to compare the distribution of actual and predicted violent crime rates.

```{r}
ggplot(crime) +
  geom_histogram(aes(x = vcr2001), fill = "blue", alpha = 0.5, binwidth = 1) +
  geom_histogram(aes(x = predicted_vcr), fill = "red", alpha = 0.5, binwidth = 1) +
  labs(title = "Overlay of Actual vs Predicted Violent Crime Rates",
       x = "Violent Crime Rate",
       y = "Density",
       caption = "Note: Blue=Actual Rates, Red=Predicted Rates") +
  theme_minimal()
```

*Question:* How do the histograms of the actual and predicted violent crime rates compare?
What do you think of the model's predictive accuracy?

*Answer:*

**Plotting Predicted vs Actual Violent Crime Rates**

```{r}
ggplot(crime, aes(x = vcr2001, y = predicted_vcr)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "blue") + # Add reference line on which actual = predicted
  scale_x_continuous(limits=c(0,150)) +
  scale_y_continuous(limits=c(0,150)) +
  labs(title = "Plot of Predicted vs Actual Violent Crime Rates",
       x = "Actual Violent Crime Rates",
       y = "Predicted Violent Crime Rates") +
  theme_minimal()
```

*Question:* If our model accounted for all the variation in violent crime rates, how would the plot of predicted vs actual violent crime rates look different?

*Answer:* 

Visualizing the Predicted versus Actual for each observation

```{r}
crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  geom_smooth(method="lm", color="red", fill=NA) +
  geom_segment(aes(xend = poverty_rate, yend = predicted_vcr), color="blue", linetype="dotted", size=0.7) +
  ggtitle("Relationship Between % Poverty and Violent Crime Rate") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty", caption="Notes: Washington D.C. census tracts, 2000") +
  theme_minimal()
```

*Question:* What do the vertical blue dotted lines represent?

*Answer:* 

**Generating Residuals**

```{r}
crime$residual_vcr <- residuals(lm_model)

```

**Density Plot of residuals**

```{r}
ggplot(crime, aes(x = residual_vcr)) +
  geom_density(color = "red") +
  labs(title = "Density Plot of Residuals", x = "Residuals", y = "Density") +
  theme_minimal()
```

*Question:* How does the distribution of residuals compare to a normal distribution?
What implications does this have for our model?

*Answer:* 

**Plotting Residuals vs Predicted Values with Reference Line at 0 (predicted = actual)**

```{r}
ggplot(crime, aes(x = predicted_vcr, y = residual_vcr)) +
  geom_point(color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") + # reference line 
  labs(title = "Residuals vs Predicted Values",
       x = "Predicted Violent Crime Rates", y = "Residuals") +
  theme_minimal()
```

*Question:* Analyze the Residual vs Predicted plot.
Are there any patterns or is the distribution random?
What does this imply about your model?

*Answer:* 

## Supplemental Code on Centering

**Centering a Variable:** The idea behind centering a variable (like poverty rate) is to adjust the data so that its mean becomes zero.
This can help in interpreting the intercept of the regression model as it becomes the expected value of the dependent variable when the independent variable is at its average value.

Recoding to center the poverty rate variable

```{r}
crime <- crime %>%
  mutate(centered_poverty_rate = poverty_rate - mean(poverty_rate, na.rm = TRUE))
```

Centering the poverty rate variable shifts the scale so that the mean is now at zero.
You can run all your same analyses.
The same rules apply, but now your interpretation of the intercept is different...check it out.

Fit linear models

```{r}
lm_original <- lm(vcr2001 ~ poverty_rate, data = crime)
lm_centered <- lm(vcr2001 ~ centered_poverty_rate, data = crime)
```

Generate the summary table. Note that you can wrap multiple models using the `list` function.

```{r}
modelsummary(list(lm_original, lm_centered),
             stars = T)

```

*Explanation:* The first model uses the original poverty rate, where the intercept indicates the expected violent crime rate when the poverty rate is zero.
The second model uses the centered poverty rate, where the intercept now represents the expected violent crime rate when the poverty rate is at its average (because the average is now centered to zero!)

## Way to Go! End of Week 2 Lab
