---
title: "SOC542 Week 9: Predictions and Marginal Effects"
author: "YJ Chae"
date: "Spring 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk_set(echo = TRUE, warning = FALSE, message = FALSE, options(scipen = 10))

library(tidyverse)
library(haven)
library(marginaleffects)
library(modelsummary)

set.seed(1234)

# Load data once for all examples
gss <- read_dta("../DATASETS/GSS/GSS2022.dta") %>%
  transmute(
    income = as.numeric(conrinc),
    age = ifelse(age < 0, NA, age),
    educ = ifelse(educ < 0, NA, educ),
    sex = factor(sex, levels = 1:2, labels = c("Male", "Female")),
    race = factor(race, levels = 1:3, labels = c("White", "Black", "Other")),
    unemp = ifelse(unemp == 1, 1, 0)
  ) %>% 
  drop_na()
```

## Introduction: The `marginaleffects` Workflow

Instead of interpreting raw coefficients (log-odds, etc.), we use the **`marginaleffects`** package to calculate quantities of interest. We categorize these into three main functional groups: **Predictions**, **Comparisons**, and **Slopes**.

---

## 1. Predictions
Predictions tell us the "absolute" expected value of our outcome for specific cases.

### `predictions()`
Useful for "What If" scenarios at specific points.
```{r predictions}
mod_linear <- lm(income ~ educ + sex + race, data = gss)

# Predicted income for every individual in the dataset
predictions(mod_linear)

# Predicted income for a Female with 16 years of education (at means of other variables)
predictions(mod_linear, newdata = datagrid(educ = 16, sex = "Female"))
```

### `avg_predictions()`
Calculates the average predicted value across your entire sample (or specific groups).
```{r avg_predictions}
mod_logit <- glm(unemp ~ age + educ + sex, data = gss, family = binomial)

# Average predicted probability of unemployment at the modal values.
avg_predictions(mod_logit)

# Average predicted probability of unemployment for Males vs. Females, using modal values for other variables.
avg_predictions(mod_logit, variables = "sex")
```

---

## 2. Comparisons
Comparisons show us the "difference" between two levels or groups (e.g., Contrast).

If the `predictions` function family pertains to predicted Y, this `comparisons` family focuses on the intuitive interpretation of your beta coefficients.

### `comparisons()`
Calculating specific differences between categories or values.
```{r comparisons}

# Difference in predicted income between Male and Female, with modal values for all other variables.
comparisons(mod_linear, variables = "sex")

# Difference in predicted income between Male and Female at 18 years of education
comparisons(mod_linear, variables = "sex", newdata = datagrid(educ = 18))
```

### `avg_comparisons()`
Summarizes the "Average Marginal Effect" across the whole sample.
```{r avg_comparisons}

# Average difference in the probability of unemployment between genders.
avg_comparisons(mod_logit, variables = "sex")
```

---

## 3. Slopes (most useful when working with quadratic forms or interactions)
Slopes (aka. marginal effects, dydx, y-prime, first derivative, etc.) tell us the `rate of change` for continuous variables.

### `slopes()`
The instantaneous effect at a specific point (useful for nonlinear models).
```{r slopes}
mod_nonlinear <- lm(income ~ age + I(age^2), data = gss)

# The "Return on Age" at 25 vs 65 (slope of the curve)
slopes(mod_nonlinear, variables = "age", newdata = datagrid(age = c(25, 65)))
```

### `avg_slopes()`
The average rate of change across all observations in your data.
```{r avg_slopes}
# Average effect of an additional year of education on probability of unemployment
avg_slopes(mod_logit, variables = "educ")
```

---

## 4. Plotting Functions
Visualizing these results is the most effective way to communicate your findings.

### `plot_predictions()`
```{r plot_predictions}

plot_predictions(mod_linear, condition = c("educ", "sex")) +
  theme_minimal() + labs(title = "Predicted Income Trend")
```

### `plot_comparisons()`
```{r plot_comparisons}

plot_comparisons(mod_logit, variables = "sex", condition = "educ") +
  theme_minimal() + labs(title = "Gender Gap in Unemployment by Education")
```

### `plot_slopes()`
```{r plot_slopes}

plot_slopes(mod_nonlinear, variables = "age", condition = "age") +
  theme_minimal() + labs(title = "How the 'Effect' of Age Changes Over Time")
```

---

## 5. Lab Exercise: Returns on Education

In this exercise, you will apply all three functional groups to investigate gender disparities in the economic returns on education.

### **The Model**
First, let's fit an interaction model to see if the impact of education on income depends on gender.
```{r lab-model}

mod_lab <- lm(income ~ educ * sex + race, data = gss)
```

### **1. Prediction (Y-hat)**
What is the absolute predicted income for a **Female** with a PhD (20 years of education)?
```{r lab-prediction}
# Use predictions() to find the value for a specific case
predictions(mod_lab, newdata = datagrid(educ = **BLANK**,
                                        sex = **BLANK**))
```

### **2. Comparison (The Gender Gap)**
Using `avg_comparisons()`, what is the **average gender gap** in income across your entire sample?
```{r lab-comparison}

# Use avg_comparisons() to find the average difference (Female - Male)
avg_comparisons(mod_lab, variables = **BLANK**)
```

### **3. Slope (The "Return" on Education)**
How much does the **"Return on Education"** (the slope) differ for a Male vs. a Female? Calculate the slope of income for each gender.
```{r lab-slope}

# Use slopes() to calculate the marginal effect of education by sex
slopes(mod_lab, variables = **BLANK**, by = **BLANK**)
```

### **4. Visualization **
Visualize the widening/narrowing gap as education increases from 12 to 20 years.
```{r lab-plot}

plot_predictions(mod_lab, condition = c("educ", "sex")) +
  labs(title = "The Gender Gap in Returns on Education",
       subtitle = "Predicted Income with 95% Confidence Intervals",
       x = "Years of Education",
       y = "Predicted Income ($)") +
  theme_minimal()
```
