---
title: "Week 10 - Count Outcomes"
author: "YJ Chae"
date: "2026-04-06"
output:
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(modelsummary)
library(marginaleffects)
library(MASS)
library(pscl)
library(haven)
library(rstanarm)
library(brms)
library(broom.mixed)

```

# 1. Introduction
This lab explores count outcomes using the number of children (`childs`) as our primary dependent variable. We compare Poisson, Negative Binomial, Zero-Inflated, and Bayesian approaches.

# 2. Data Preparation (GSS 2022)
```{r data}
gss_raw <- read_dta("DATASETS/GSS/GSS2022.dta")

gss <- gss_raw %>%
  transmute( 
    childs = ifelse(childs < 0, NA, as.numeric(childs)),
    age = ifelse(age < 0, NA, as.numeric(age)),
    educ = ifelse(educ < 0, NA, as.numeric(educ)),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    marital = factor(marital, levels = 1:5, labels = c("Married", "Widowed", "Divorced", "Separated", "Never Married"))
  ) %>%
  drop_na()

```

Let's first take a look at the distribution of the number of children in GSS 2022.

```{r vis}
ggplot(gss, aes(x = childs)) +
  geom_bar(fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Number of Children",
       subtitle = "Source: GSS 2022",
       x = "Number of Children",
       y = "Frequency")

```

# 3. Poisson Regression
The Poisson distribution is used for modeling the number of times an event occurs in a fixed interval. The probability of observing a count $y$ is:

$$P(Y=y) = \frac{e^{-\mu} \mu^y}{y!}$$

where $\mu$ is the expected count (the rate). In Poisson regression, we model the natural log of this rate:

$$\ln(\mu) = \beta_0 + \beta_1X_1 + \dots + \beta_kX_k$$

Exponentiating the coefficients ($e^\beta$) gives **Incident Rate Ratios (IRR)**, which tell us how the rate changes for a one-unit increase in $X$. Poisson assumes Mean = Variance.

```{r poisson}
m_poi <- glm(childs ~ age + educ + sex + marital,
             data = gss,
             family = "poisson")

modelsummary(list("Log-Count" = m_poi, "IRR" = m_poi),
             exponentiate = c(FALSE, TRUE),
             stars = TRUE)
```

### 3.1 Overdispersion Check
```{r poi_check}
paste("Deviance/DF Ratio:", round(m_poi$deviance / m_poi$df.residual, 2))

```

# 4. Negative Binomial Regression
The Negative Binomial distribution is an extension of the Poisson that allows for **overdispersion** (Variance > Mean). We use this when the Poisson's assumption of Mean = Variance is violated, often indicated by a Deviance/DF ratio > 1.


```{r negbin}
m_nb <- glm.nb(childs ~ age + educ + sex + marital, data = gss)

modelsummary(list("Poisson" = m_poi, "NegBin" = m_nb),
             exponentiate = TRUE,
             metrics = c("AIC","BIC"))
```

# 5. Zero-Inflated Poisson (ZIP)
Zero-inflated models are used when there are "extra" zeros beyond what a standard count distribution would predict. The ZIP model combines a **logit model** (to predict the probability of being a structural zero, $\pi$) and a **Poisson model** (to predict the count, $\mu$):


```{r zip}
# Note the pipe (|), separating two sets of coefficients:
# Formula: DV ~ IDV for count | IDV for zero

m_zip <- zeroinfl(childs ~ age + educ + sex | age + educ + sex,
                  data = gss,
                  dist = "poisson")
summary(m_zip)

modelsummary(m_zip, exponentiate = T, stars = T)

```

# 6. Bayesian Count Models
Bayesian models allow for robust uncertainty estimation and posterior predictive checks.

```{r bayes}
m_stan_poi <- stan_glm(childs ~ age + educ + sex, data = gss,
                       family = "poisson",
                       refresh = 0,
                       seed = 1234)

m_stan_nb <- stan_glm(childs ~ age + educ + sex,
                      data = gss,
                      family = "neg_binomial_2",
                      refresh = 0,
                      seed = 1234)

## To use modelsummary with Bayes NB models
{
  tidied_neg = tidy(m_stan_nb, conf.int = T)
  glance_neg = glance(m_stan_nb)
  msummary_out = list(tidy = tidied_neg,
                      glance = glance_neg) 
  class(msummary_out) = 'modelsummary_list'
}

modelsummary(list("Bayes Poisson" = m_stan_poi, "Bayes NegBin" = msummary_out),
             statistic = "conf.int",
             exponentiate = TRUE)

```

### 6.1 Posterior Predictive Checks
```{r pp_check}
pp_check(m_stan_poi) + ggtitle("Posterior Check: Children (Poisson)")
pp_check(m_stan_nb) + ggtitle("Posterior Check: Children (NegBin)")

```

# 7. Model Comparison: Predicted vs. Actual Distributions
To see how well our models handle the distribution (especially the zeros), we can compare the predicted outcomes from each model against the actual sample distribution.

```{r comparison_plot}
# 1. Generate Predicted Counts (Expected Values) for each observation
# We use posterior_epred for Bayesian models to get the expected value
gss_comp <- gss %>%
  mutate(
    Actual = childs,
    Poisson = predict(m_poi, type = "response"),
    NegBin = predict(m_nb, type = "response"),
    ZIP = predict(m_zip, type = "response"),
    Bayes_Poisson = colMeans(posterior_epred(m_stan_poi)),
    Bayes_NegBin = colMeans(posterior_epred(m_stan_nb))
  )

# 2. Reshape to long format for ggplot
plot_data <- gss_comp %>%
  pivot_longer(
    cols = c(Actual, Poisson, NegBin, ZIP, Bayes_Poisson, Bayes_NegBin),
    names_to = "Model",
    values_to = "Value"
  )

# 3. Density Plot Comparison
ggplot(plot_data, aes(x = Value, fill = Model, color = Model)) +
  geom_density(alpha = 0.1) +
  theme_minimal() +
  scale_x_continuous(limits = c(0, 8), breaks = 0:8) +
  labs(title = "Distribution Comparison: Actual vs. Predicted Counts",
       subtitle = "Including Frequentist and Bayesian Estimates",
       x = "Number of Children (Expected Value)",
       y = "Density")

```


# 8. Summary
1. **Poisson**: Baseline model for count data.
2. **Negative Binomial**: Handles overdispersion (Variance > Mean). If not sure, use this model!
3. **Zero-Inflated**: Handles excess zeros via a logit + Poisson.
4. **Bayesian**: Robust uncertainty estimation. Use `pp_check` and see posterior predictive distributions.
