---
title: "Week 6 - Interaction Terms"
author: "YJ Chae"
date: "03-02-2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(haven)
library(marginaleffects)
library(modelsummary)
options(scipen = 999)
```

# Data Preparation
```{r data}
# Load and clean data (including log_inc)
gss_raw <- read_dta("DATASETS/GSS/GSS2022.dta")

gss <- gss_raw %>%
  select(conrinc, hrs1, wrkslf, degree, sex, race, age, educ) %>%
  mutate(
    income = as.numeric(conrinc),
    log_inc = log(income + 1), # Add 1 to avoid log(0)
    age = ifelse(age < 0, NA, age),
    workhrs = ifelse(hrs1 < 0 | hrs1 >= 89, NA, hrs1),
    educ = ifelse(educ < 0, NA, educ),
    female = ifelse(sex == 2, 1, 0),
    self_employed = ifelse(wrkslf == 1, 1, 0),
    college = ifelse(degree >= 3, 1, 0),
    race = factor(race, labels = c("White", "Black", "Other")),
    degree = factor(degree, levels = c(1,0,2,3,4), labels = c("HS", "Less_HS", "Assoc", "Bach", "Grad"))
  )

```

# 0. Recap: Log-Linear Models
Before we dive into interactions, let's recap the log-income model from Week 5. When we log-transform the dependent variable, we are interested in multiplicative changes (percentages) rather than additive changes. It's important to understand how to interpret coefficients correctly!

```{r recap-log-model}
# Week 5 Log Model
m_log <- lm(log_inc ~ degree + workhrs + female + race, data = gss)

# Comparison Table: Standard vs Exponentiated
modelsummary(list("Log-Scale" = m_log, "Exponentiated" = m_log),
             exponentiate = c(FALSE, TRUE),
             stars = TRUE,
             gof_omit = ".*",
             title = "Week 5 Recap: Log-Income Model")
```

When your dependent variable is somehow log-transformed, the `exponentiate` argument in `modelsummary` helps you interpret the coefficients intuitively.

---

# 1. Understanding Interaction Logic
An interaction occurs when the effect of one variable **depends on the level of another**.

### The Math of Interaction
Consider a model where income depends on education ($Edu$) and being female ($Female$):
$$\hat{Y} = \beta_0 + \beta_1 Edu + \beta_2 Female + \beta_3 Edu \cdot Female$$

To see how the effect of education ($Edu$) changes for women, we factor the equation:
$$\hat{Y} = \beta_0 + (\beta_1 + \beta_3 Female) Edu + \beta_2 Female$$

*   If $Female = 0$ (Male), the effect of education is simply **$\beta_1$**.
*   If $Female = 1$ (Female), the effect of education is **$\beta_1 + \beta_3$**.

$\beta_3$ tells us exactly how much the "return on education" changes when you are female.

Finally, when interacting two variables, the coefficient for each main term tells you its effect when the *other* variable equals zero. Let's figure this out again by factoring the terms!


# 2. Categorical X Categorical Interactions
Let's look at how the female disadvantage in income might differ by self-employment status.

```{r categorical-interaction}
# 1. Additive Model: Effect of being female is assumed constant
m1_add <- lm(income ~ college + female, data = gss)

# 2. Interaction Model: Effect of being female can vary by college, or... the other way around?
m1_int <- lm(income ~ college * female, data = gss)

modelsummary(list("Additive" = m1_add, "Interaction" = m1_int),
             estimate = "{estimate}{stars}",
             fmt = 0,
             gof_omit = ".*")

```

**Question:** In the interaction model, what does the coefficient for `college x female` represent? How about the main terms? Using the factoring logic above, let's find out the total effects of each college education and being female.

**Question 2:** What is the **direction** of the interaction terms? Does `college` affect the coefficient of `female`, or the other way around, or is it bidirectional?


*Your Answer:* 

Great. We can calculate the effect size for college-graduate women by hand, but it doesn't give us any p-value, right? We can conduct a hypothesis test whether this coefficient is different from zero using the `marginaleffects` package.

```{r marginaleffects-1}
avg_comparisons(m1_int, variables = "college", by="female")
plot_comparisons(m1_int, variables = "college", by="female") +
  ggtitle("The Effect of College Education by Gender") +
  theme_minimal()

```

This marginaleffects calculation and plot show the effect of college education on income by gender.

---

# 3. Categorical X Continuous Interactions
Does the "return" on each additional hour of work differ for those with a college degree?

```{r cat-con-interaction}
m2_add <- lm(income ~ workhrs + college, data = gss)
m2_int <- lm(income ~ workhrs * college, data = gss)

modelsummary(list(m2_add,m2_int), estimate = "{estimate}{stars}", fmt = 0, gof_omit = ".*")

```

**Question:** If the interaction term `workhrs:college` is positive, what does that tell us about the value of an extra hour of work for college graduates vs. non-graduates? 

**Question 2:** In this case, how does a **one unit increase** in work hours differ from one unit increase in `college`?

*Your Answer:* 

```{r marginaleffects-2}
avg_comparisons(m2_int, variables = "workhrs", by="college")
plot_comparisons(m2_int, variables = "workhrs", by="college") +
  ggtitle("The Effect of Work Hours by College Education") +
  theme_minimal()

```

---

# 4. Continuous X Continuous Interactions
Does the effect of years of education on income depend on how many hours you work?

```{r con-con-interaction}
# Task: Create a model interacting education and work hours
m3_add <- lm(income ~ educ + workhrs, data = gss)
m3_int <- lm(income ~ educ * workhrs, data = gss)

modelsummary(list(m3_add,m3_int), estimate = "{estimate}{stars}", fmt = 0, gof_omit = ".*")

```

**Question:** What does the coefficients for the main terms, `workhrs` and `educ`, mean in this case? Why is the intercept so negative?

*Your Answer:* 

```{r marginaleffects-3}
avg_comparisons(m3_int, variables = "workhrs", by="educ")
plot_comparisons(m3_int, variables = "workhrs", by="educ") +
  ggtitle("The Effect of Work Hours by Years of Education") +
  theme_minimal()

```


# 5. Visualizing Continuous X Continuous Interactions using Bin
Interactions between two continuous variables can be difficult to interpret from coefficients alone. We can use the `marginaleffects` package to visualize how the "slope" of one variable (e.g., `workhrs`) changes at different model values of the other (e.g., `educ` at 4, 12, and 16 hours).

```{r plot-con-con}
# Plotting the interaction from Section 4
# We use 'condition' to specify which values of workhrs to highlight
plot_predictions(m3_int, condition = list("workhrs", "educ" = c(4, 12, 16))) +
  labs(title = "Effect of Work Hours at Different Model Values of Years of Education",
       y = "Predicted Income ($)",
       x = "Work Hours",
       color = "Years of Education",
       fill = "Years of Education",
       ) +
  theme_minimal()
```


### Squared Terms as "self-interactions"
Using this same logic, we can see that the squared terms (polynomials) we used in Week 5 are technically a variable **interacting with itself**.

Think of the equation for work experience:
$$\hat{Y} = \beta_0 + \beta_1Exp + \beta_2Exp^2$$

By factoring out $Exp$:
$$\hat{Y} = \beta_0 + (\beta_1 + \beta_2 \cdot Exp) \cdot Exp$$

The "effect" of experience is $(\beta_1 + \beta_2 \cdot Exp)$. Just like with the female/education interaction, the effect of an extra year of experience depends on the value of another variable—in this case, your current level of experience! This enables modeling of a curvillinear relationship--a quadratic polynomial.

Great job, everyone. Let's keep it up!
