---
title: "Week 8 - Regression with a Binary Outcome"
author: "YJ Chae"
date: "03-23-2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(haven)
library(modelsummary)
library(rstanarm)
library(broom.mixed)
library(DescTools)
library(marginaleffects)

set.seed(1234)
```

# 1. Data Preparation

We'll investigate factors associated with support for marijuana legalization using the 2022 GSS.

```{r data-prep}
gss_raw <- read_dta("DATASETS/GSS/GSS2022.dta")

gss <- gss_raw %>%
  select(grassnv, polviews, sex, race, age, educ, wtssnrps) %>%
  mutate(
    # Target: 1 = Legal, 0 = Not Legal
    legal_grass = case_when(
      grassnv == 1 ~ 1,
      grassnv == 2 ~ 0,
      TRUE ~ NA
    ),
    conservatism = polviews, # 1 (Lib) to 7 (Cons)
    age = ifelse(age > 88, NA, age),
    female = ifelse(sex == 2, 1, 0),
    race = factor(race, labels = c("White", "Black", "Other"))
  ) %>%
  drop_na(legal_grass, conservatism, age, female, race)
```

---

# 2. Linear Probability Model (LPM)

The LPM is a standard OLS regression on a binary outcome. It estimates the change in the **probability** of the outcome.

```{r lpm}
mod_lpm <- lm(legal_grass ~ conservatism + age + female + race, 
              data = gss)

# The main issue: predicted probabilities can exceed [0, 1]
p_lpm <- predict(mod_lpm)
summary(p_lpm)
```

---

# 3. Logistic Regression (Logit)

In OLS, we model the outcome $Y$ directly, meaning that we assume the dependent variable ranges from $-\infty$ to $\infty$ while the true range is [0, 1]. Instead, we must model the **probability** $p$ that $Y=1$. To ensure our predictions stay between 0 and 1, we use a *link function*.

### The Mathematics of Logit

The relationship between probability ($p$), odds, and log-odds (logit) is the foundation of this model:

1.  **Probability ($p$):** The likelihood of the event occurring (ranging from 0 to 1).
2.  **Odds:** The ratio of the probability of the event occurring to that not occurring:
    $$\text{Odds} = \frac{p}{1-p}$$
3.  **Log-odds (Logit):** The natural logarithm of the odds. This transforms the 0–1 probability into an unbounded line $(-\infty, \infty)$, which we can then model linearly:
    $$\text{logit}(p) = \ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \dots + \beta_k X_k$$
    
4.  **The Logistic Function:** To get back to a probability from our linear equation, we use the inverse-logit (logistic) function. The logistic function is:
$$\text{Logistic}{(x)} = \frac{1}{1 + e^{-x}}$$.

That looks unfamiliar, but if we input our dependent variable, $$\text{Logistic}{(\text{logit(p)})} = p$$.

Therefore, $$p = \text{Logistic}{(\text{logit}{(p)})} = \text{Logistic}\left(\ln\left(\frac{p}{1-p}\right)\right) = \frac{1}{1 + e^{-\ln\left(\frac{p}{1-p}\right)}} = \frac{1}{1 + e^{-(\beta_0 + \sum \beta_i X_i)}}$$

**Key Insight:** While the relationship between $X$ and the **logit** is linear and unbounded, the relationship between $X$ and the **probability** is S-shaped and bounded between 0 and 1. This ensures our predictions never fall below 0 or rise above 1.

Before getting into the actual estimation process, let's think about what it means to explain a social concept with an equation. Do you think this works, or not? Why?
While a mathematical equation might seem too rigid for the complexity of social life, it serves as a measurement of the quantity of interest. By reducing a phenomenon (e.g., support for legalization) into a set of coefficients, we aren't claiming to capture every nuance of human experience. Instead, we are testing whether we can detect a meaningful pattern within individual variation. 


### 3a. Estimation
We use `glm()` with `family = "binomial"`.

```{r logit}
mod_logit <- glm(legal_grass ~ conservatism + age + female + race, 
                 family = binomial(link = "logit"), # You can simply use family = binomial().
                 data = gss)

summary(mod_logit)

```

```{r logit-modelsummary}
# Comparisons
modelsummary(list("LPM (OLS)" = mod_lpm, "Logit (MLE)" = mod_logit),
             stars = TRUE, 
             gof_omit = "Log|IC|F|RMSE")

```

### 3b. Interpretation: Odds Ratios
Logit coefficients are in "log-odds." To make them intuitive, we exponentiate them into **Odds Ratios (OR)**.

*   $OR > 1$: Increase in odds
*   $OR < 1$: Decrease in odds

Note this boilerplate language to interpret ORs: "One-unit increase in conservatism is associated with 44.4% decrease (i.e., 1-0.556) in the odds of support for legalization.

We can also compare the linear line of log-odds vs. the S-curve (p). Let's use our blackboard.

```{r odds-ratios}
modelsummary(
  list("LPM (OLS)" = mod_lpm, "Logit (Log-Odds)" = mod_logit, "Logit (ORs)" = mod_logit),
  exponentiate = c(F,F,T), # Note: DO not exponentiate the first two models, and exponentiate the third one.
  statistic = "conf.int", 
  stars = TRUE, 
  gof_omit = "Log|IC|F|RMSE",
  title = "Models with a binary outcome")

```

### 3c. Marginal Effects (Average Marginal Effects)

While odds ratios (ORs) are helpful, they have a few known issues, which we won't discuss here. Also, social scientists often prefer to discuss the effect of a variable in terms of **probability points** (similar to the LPM). Average Marginal Effects (AME) provide the average change in the predicted probability across all observations in the sample.

```{r marginal-effects}
# Average Marginal Effects (AME)
# We use get_slopes to get the underlying data for modelsummary
mod_logit_ame <- avg_comparisons(mod_logit)

# Modelsummary shows probabilities, not log-odds
mod_logit_ame

```

We can also visualize how the predicted probability changes across specific variables (Predicted Marginal Means).

```{r marginal-plots}
# Predicted probability across the range of conservatism
plot_predictions(mod_logit, condition = "conservatism") +
  theme_minimal() +
  labs(title = "Effect of Conservatism on Support for Legalization",
       y = "Predicted Probability", x = "Conservatism (1-7)")

# Predicted probability by race and sex
plot_predictions(mod_logit, condition = c("race", "female")) +
  theme_minimal() +
  labs(title = "Predicted Probability by Race and Gender",
       y = "Predicted Probability")
```

---

# 4. Model Evaluation

### 4a. Predicted Probabilities
Unlike the LPM, Logit predictions are bounded.

```{r logit-pred}
gss$pred_logit <- predict(mod_logit, type = "response")

ggplot(gss, aes(x = pred_logit)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Predicted Probabilities (Logit)",
       x = "Probability of Supporting Legalization") +
  theme_minimal()

```

### 4b. Pseudo R-Squared
Since OLS $R^2$ doesn't exist for Logit, we use "pseudo" $R^2$ measures. A common standard is McFadden's $R^2$.

```{r pseudo-r2}
# Using DescTools for multiple pseudo-R2 metrics
DescTools::PseudoR2(mod_logit, which = c("McFadden", "CoxSnell", "Nagelkerke"))

```

---

# 5. Bayesian Binary Regression

Finally, we estimate the model using Bayesian methods for more robust uncertainty estimates.

```{r bayes-logit}
mod_bayes <- stan_glm(legal_grass ~ conservatism + age + female + race, 
                      family = binomial(link = "logit"), 
                      data = gss,
                      refresh = 0, seed = 1234)

# Posterior Predictive Check
pp_check(mod_bayes) +
  labs(title = "Bayesian Posterior Predictive Check")
```

---

# Summary for regression with a binary outcome:
1. Use **Logit** (`glm`) for empirically rigorous binary modeling.
2. Interpret Logit using **Odds Ratios** (`exponentiate = TRUE` in `modelsummary` package).
3. To interpret coefficients intuitively, estimate average marginal effects (AMEs) using `marginaleffects` package.
4. Evaluate performance using **Log-Likelihood** and **Pseudo R^2**.
