---
title: '542: Week 2 - Regression Review'
author: "Fred Traylor, Lab TA"
date: "1/31/2022"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)

library(tidyverse)
library(gapminder)
library(stargazer)
```


# Loading and Visualizing Our Variables

We're going to use the `gapminder` package to load up our data. The Gapminder project has data on a wide variety of indicators, for most countries of the world, and for many decades. This provides a great opportunity for us to analyze development. 

For this week, we're going to use just a sample of data from the United States of America. This will make it easier to see all our points. 

```{r data-load}
usadata <- gapminder::gapminder_unfiltered %>% 
  filter(country == "United States") %>% 
  select(-country, -continent)
head(usadata) # shows the top of the table
```

Looking at the data information with `?gapminder`, we can see that there are six variables.

1.  country = Country Name (removed from our dataset on line 33)
2.  continent = Continent Name (removed from our dataset on line 35)
3.  year = Year, ranges from 1952 to 2007
4.  lifeExp = Life expectancy at birth, in years
5.  pop = population
6.  gdpPercap = GDP per capita, USD, inflation-adjusted

Let's take a look at these data. And especially, let's look at the relationship between year and life expectancy.

```{r data-viz}
ggplot(usadata, aes(x = year,y = lifeExp)) +
  geom_line() +
  geom_point()
```

# OLS Regression
OLS regression in R uses the `lm()` function, short for "Linear Model." Below, let's regress life expectancy on year to see how life expectancy has increased over time. 

We'll also use two ways to display the results, `summary()` and `stargazer::stargazer()`. 

```{r regression}
mod1 <- lm(lifeExp ~ year,   # Regress life expectancy on year
           data = usadata)   # Using data from usadata 

summary(mod1)
```

From these results, we can see that life expectancy in the USA increased at a rate of 0.182 additional years of expected life each year, starting at -287.415 years in the year zero. (Of course, we didn't start at year 0; our data start in year 1952.)

Looking back at our summary table, our R-squared is very high (.981), suggesting that around 98% of the variations in life expectancy can be explained by the passage of time.

We can use the `stargazer` package to create cleaner looking regression tables for publications. Additional arguments can be supplied to customize the appearance of the table.

```{r regression-output}
stargazer(mod1, 
          type = "text")
```

# Statistical Significance of Coefficients

There are three ways we can see if the coefficient in our model (year) is significant.
1. P-value: The p-value is very low. We can see this in the first print-out that it is very close to zero. We can also see the stars attached to it in both the `summary()` and `stargazer()` print-outs. 
2. T-value: The t-value is 53.84. The t-statistic is calculated as the ratio of the standard error to the coefficient. In this case, it is $.182 / .003 = 53.84$. A value larger than 2 (or less than -2) points to a p-value less than .05, and it only increases in significance from there. 
3. Confidence interval: While we don't get it directly in our print-outs, we can tell `stargazer()` to display it in our table. 

```{r}
stargazer(mod1,
          ci = T,
          ci.level = .99, # 99%
          type = "text")
```

Because our confidence interval doesn't cross zero, we know that it is significant at the $\alpha = .01$ level. 

# Predictions
After running a regression, we often want to use it to predict a specific value. Recalling that our regression equation is $y = a + bx$, we can substitute our intercept for $a$ and our year coefficient for $b$, giving us the new equation, $lifeExp = -287.415 + 0.182*year$. Using this equation, we can plug in any point we want for the value of year and get a predicted value for life expectancy. 

*NOTE: These equations are bookended with dollar signs so that, in the knitted version of this document, they'll be converted to a readable version of an equation. You can hover your mouse over any part of the equation, and it'll show you what it would look like in the document.*

While it is possible to type in the coefficients and intercept ourselves, any changes to our model (in data or formula) will mean having to retype it all again. 

In this case, we can directly access the coefficients in the `mod1` object. We can then use indexing to access the relevant elements. Note that some incorrect text naming is left over the first time. We can escape this by using `[[]]` to select the contents of each value.


```{r predict-byhand}
new_year <- 1970

-287.414516 + 0.182311*new_year

mod1$coefficients

mod1$coefficients[1] + mod1$coefficients[2]*new_year

mod1$coefficients[1][[1]] + mod1$coefficients[2][[1]]*new_year # better

```

R also has the `predict()` function, which does the same thing. Just note, though, that the new year must be in the form of a data frame. This means we can easily calculate multiple predicted values at once.
```{r predict-function}
new_year = data.frame(year = 1970)

pred <- predict(mod1, new_year)
print(pred)
```

# Residuals 
How far is this predicted point from the actual value for 1970, though? Looking at our dataset, we can see that it was actually 70.92, a little bit away from the predicted value above. We can calculate exactly how far it is by subtracting the predicted value from the actual value.

```{r resid-byhand}

realval <- usadata %>% 
  select(year, lifeExp) %>% 
  filter(year == 1970)
head(realval)

pred - realval$lifeExp
```

Like with `predict()`, R has another built-in function to shorten this process for us, called `resid()` or `residuals()`. Let's use that function, and our `predict()` function from earlier, to add the predicted and residual values onto our original dataframe.
```{r add-predict-resid}
usadata$predicted <- predict(mod1)
usadata$resid <- residuals(mod1)

usadata %>% 
  select(year, lifeExp, predicted, resid) %>% 
  head()

```
Note that `lifeExp - predicted = resid`.

## Graphing Residuals 

A first look at the residuals warrants seeing if there are any values that are poorly predicted. We can graph a histogram of residuals to see this.
```{r resid-hist}
ggplot(data = usadata, aes(x = resid)) +
  geom_histogram(bins=20)
```
This shows us that, while most residuals are very close to zero, there are still a few that are further away. 

Since residuals can also be negative, let's also graph a histogram of the absolute values of the residuals.
```{r resid-hist-abs}
ggplot(data = usadata, aes(x = abs(resid))) +
  geom_histogram(bins=20) 
```

A handy graph in inspecting our regression model is the residual plot. Our x-axis here is the predicted values of life expectancy, and our y-axis is the residuals. 

```{r resid-plot}

ggplot(usadata,
       aes(x = predicted,
           y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  ggtitle("Residual Plot") +
  theme_classic()

```

Lastly, let's plot this graph side-by-side with our original graph. We're going to use the `plot_grid()` function, contained in the `cowplot` package. Create each graph separately, save each as its own object, then run `cowplot::plot_grid(graph1, graph2)`.


```{r compare-graphs}
origplot <- ggplot(usadata,
                   aes(x = year,
                       y = lifeExp)) +
  geom_point() +
  geom_abline(intercept = mod1$coefficients[1], 
              slope = mod1$coefficients[2],
              linetype = "dashed",
              color = "red", size = 1) +
  ggtitle("Life Expectancy vs. Year") +
  theme_classic()

residplot <- ggplot(usadata,
                    aes(x = predicted,
                        y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "red", size = 1) +
  ggtitle("Residual vs Predicted Plot") +
  theme_classic()

if(!require(cowplot))          # If cowplot package isn't intalled...
  install.packages("cowplot")  # Install it

cowplot::plot_grid(origplot, residplot)
```

Looking at it, we can see that the residuals increase and decrease around the same places in the original plot and in the residual plot.  

