---
title: "Week 7 - Interaction Terms"
author: "Brent Hoagland, Lab TA"
date: "3/3/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
options(scipen = 999)

library(tidyverse)
library(haven)

library(modelsummary)
library(broom.mixed)
  options(modelsummary_get = "broom") 

set.seed(1234) # Optional, for reproducibility
```

# Data Recoding & Wrangling

Let's play with the GSS again. This week, we don't have a specific hypothesis about what leads to income or occupational prestige, but instead we'll be playing around with various relationships between sets of variables.

```{r gss-loading}
# Write the code or change your working directory to properly load the GSS2022.dta file and assign it to the object gss2022
gss2022 <- read_dta("../DATASETS/GSS/GSS2022.dta")

gss <- gss2022 %>% 
  select(conrinc,                 # DV: Inflation-adjusted personal income 
         hrs1, wrkslf, prestg10, degree,  
         # IVs: career variables (number of hours worked last week, r self-emp or works for somebody,
         # r's occupational prestige score, r's highest degree )
         sex, race, age, educ, partyid) # IVs: some demographic information

gss <- haven::zap_label(gss) # This function takes away any long labels that NORC might have included in the dataset; it just helps for readability for view(gss)


# This time I've done all the recoding
gss <- gss %>%
  mutate(
    # Rename 'conrinc' variable to a numeric variable called 'income'
    income = as.numeric(conrinc),
    
    # Log transform of income
    logincome = log(income),
    
    # Recode 'age' by replacing -100, -99, -98 with NA
    age = ifelse(age %in% c(-100, -99, -98), NA, age),
    
    # Recoding a new variable based on age
    workexp = age - 18,
    
    # Recode 'hrs1' into 'workhrs' by replacing certain values with NA
    workhrs = as.numeric(case_when(
      hrs1 >= 89 ~ NA,      # Recoding to remove possible outliers
      hrs1 <= -97 ~ NA,     # Recoding all negative values to NA
      TRUE ~ hrs1)),         # For all remaining values of hrs1, keep the original value
    
    # Recode 'wrkslf' into 'self_employed' with descriptive labels
    wrkslf = factor(wrkslf,
                    levels = c(1, 2),
                    labels = c("Self-employed", "Works for someone")),
    
    # Create a dummary variable for self_employed
    self_employed = ifelse(wrkslf == "Self-employed", 1, 0),
    
    # Recoding 'prestg10' into a numeric variable called 'pristige'
    prestige = as.numeric(prestg10),
    
    # Recode 'degree' variable into a factor with labels
    degree = factor(degree,
                    levels = c(0, 1, 2, 3, 4),
                    labels = c("Less_HS", "HS", "Assoc", "Bach", "Grad")),
    
    # Recode 'sex' variable into a factor variable
    sex = factor(sex,
                 levels = c(1, 2),
                 labels = c("Male", "Female")),
    
    # Create a dummy variable for female
    female = ifelse(sex=="Female", 1, 0),
    
    # Recode 'race' variable into a factor with descriptive labels
    race = factor(race, levels = c(1, 2, 3),
                  labels = c("White", "Black", "Other")),
    
    # Recode 'partyid' variable into political affiliation categories
    partyid = case_when(
      partyid %in% c(0:2) ~ "Dem",
      partyid %in% c(4:6) ~ "Rep",
      partyid %in% c(3, 7) ~ "Other"
    ),
    
    # Recode 'educ' for missing values; leaving as continuous 
    educ = ifelse(educ %in% c(-99, -98), NA, educ)
  ) %>%
  
  select(-hrs1, -conrinc, -prestg10) %>%
  
  drop_na()

```

## Descriptive Statistics

As always, let's do some descriptive statistics. One thing to note is that I created dummy variables for `female`, indicating `1` whether the respondent is female and `0` if the respondent is male , and `self_employed`, indicating `1` whether they are self-employed (as opposed to working for somebody else `0`).

```{r desc-tables}

datasummary_skim(gss,
                 type = "numeric",
                 fmt = 2, # Show 2 decimal places 
                 histogram = F,
                 title = "Sample Descriptive Statistics: Continuous Variables",
                 notes = "Data: 2022 General Social Survey",
                 output = "data.frame")

datasummary_skim(gss, 
                 type = "categorical",
                 title = "Sample Descriptive Statistics: Categorical Variables",
                 notes = "Data: 2022 General Social Survey",
                 output = "data.frame")

```

**Question:** What do you notice about the numeric dummy variables for `female` and `self_employed` and their corresponding categorical/factor variable?

*Answer:* The mean for the dummy variables (0.50 and 0.12) matches the percent column for the `Female` 49.8% and `Self-employed` 12.1% response values.

# Interactions between different variable types ("X" indicates interaction)

## Dichotomous Variable (0 or 1) X Dichotomous Variable

To start, let's investigate the relationship between a person's income (in absolute dollars) and their years of education, their status as self-employed, and their sex.

I created two models below:

1.  The first is an "additive" model, where the three independent variables act independently on income.
2.  The second is a "multiplicative" model, which includes a term dependent on the status of self-employment and sex together.

```{r dum-dum}
dum_base <- lm(income ~ educ + self_employed + female,
                   data = gss)
dum_int <- lm(income ~ educ + self_employed * female, # the variable * variable indicates an interaction term
                  data = gss)

modelsummary(list("Additive" = dum_base, "Multiplicative" = dum_int),
             estimate = "{estimate}{stars}",
             fmt = 0,
             statistic = NULL,
             gof_omit = "IC|Log|alg|pss|F|RMSE",
             notes = "Notes: * p < 0.05, ** p < 0.01, *** p < 0.001",
             output="huxtable")

```

In the first model, we see that each year of education has a significant effect on income, increasing it by \$5,212 for each additional year on average. Additionally, women earned about \$17,085 less than men all else equal, while self-employment individuals made \$1,117 more holding all else equal.

**Question:** Who does the constant/intercept represent in the first model?

*Answer:* Men who work for someone else with 0 years of education.

Turning now to the second model, we must change how we interpret our table. Let's start with the interaction term: Self-employed women earned \$2,190 less income than everyone else. But who is that everyone else, and what are their values? The term `female` still denotes all women, both self- and other-employed. And the term `selfemp` represents all self-employed individuals, male and female.

Here's another way to think about it. let's create a new variable that is a cross of `female` and `selfemp`. Then, let's use that new variable in a regression and compare the output to the other.

```{r dum-dum-cross}

gss <- gss %>% 
  mutate(female_selfemp = case_when(
    female == 1 & self_employed == 1 ~ "Female, Self Emp",
    female == 1 & self_employed == 0 ~ "Female, Not Self Emp",
    female == 0 & self_employed == 1 ~ "Male, Self Emp",
    female == 0 & self_employed == 0 ~ "0 Male, Not Self Emp" 
          # putting 0 so this row is 'dropped' in regression
  ))

ftable(gss$female, gss$self_employed, gss$female_selfemp) # Double-checking the coding

dum_cross <- lm(income ~ educ + female_selfemp,
                  data = gss)

modelsummary(list("Interactive" = dum_int, "Cross" = dum_cross),
             estimate = "{estimate}{stars}",
             fmt = 1,
             statistic = NULL,
             gof_omit = "IC|Log|alg|pss|F|RMSE",
             notes = "Notes: * p < 0.05, ** p < 0.01, *** p < 0.001",
             output="huxtable")

```

Looking at the output, we can see a couple of numbers that standout. Working our way down:

1.  Education: The coefficient for education is \$5,206 in both models, indicating the consistent impact (given we specified our model this way) of each additional year of education on income, while accounting for self-employment status and sex.

2. Interpreting the Intercept (constant in this table): The coefficients match and represent the income for not-self-employed males with 0 education, which acts as the baseline against which the other coefficients are compared. 

Let's just focus on the model (2) to begin with where we accounted each combination of being self-employed and sex by creating a dummy variable (either 1 or 0) for each combination, which can be represented by a 2x2 table:

|        | Not Self-Employed | Self-Employed |
|:-------|:------------------|:--------------|
| Female | -\$16,841         | -\$17,130     |
| Male   | (intercept).      | +\$1,902      |

This table organizes the coefficients from model (2) for each combination of self-employment status and sex. By understanding model (2), we can back into model (1) and fill in our understanding of the interaction term. Here's what we observe:

3.  Female Income Penalty:  The -$16,841 coefficient in model (2) quantifies the income penalty for females versus males among those not self-employed. This figure corresponds to the female coefficient in model (1), indicating an income reduction for females relative to the reference group of not-self-employed males.

4.  Self-Employed Effect: Self-employed males earn +$1,902 more than the reference group. In contrast, self-employed females experience a -$17,130 income difference from the same reference group.

**Question:** So, does the effect of being self-employed have the same consequence for females as it does for males? That is to ask, do self-employed females make +$1,902 more than employed females (the difference between 17,130 and 16,841)?

*Answer (YES or NO):* ~ NO ~ 

5. Interpreting Self-Employed Females' Income: The -$17,130 coefficient for self-employed females in model (2) represents the combined effect of sex and self-employment status. It's not just the sum of the female penalty (-$16,841) and the self-employment income increment (+$1,901). We need to consider the main effect of female (average difference between sexes), the main effect of being self-employed, but also how being female AND self-employed combine uniquely (aka an interaction effect).

6. Interaction Effect: Model (1) suggests that self-employment corresponds to a +$1,902 income increase, and the average difference between females and males corresponds to a -$16,841 decrease against females. The interaction term self_employed:female, with a coefficient of -$2,190, implies an additional income reduction for self-employed females. The interaction effect captures the compounding disadvantage of self-employment AND female, reflected in the sum: $1,907 (selfemp) - $16,841 (female) - $2,190 (self_employed:female) ≈ -$17,130.

**Trick Question:** Does being self-employed increase someone's income? 

*Answer* It depends! Self-employment means different things/has different consequences/moderates the effect of being self-employed for males versus females. While self-employment is associated with an income increase (+$1,907) for males, it does not confer the same benefit to females, who actually face a compounded income penalty due to the interaction between sex and self-employment status (-$2,190 interaction effect + $1,907 main effect = -$288).

Wow, way to go. Interactions terms open up new possibilities to explore a different kind of relationship (aka multiplicative - it can change direction and magnitude at different points) between variables as we've done with two dichotomous variables (female x self-employed). However, with more than a simple 2X2 set of categories, it gets...well as you can probably expect...increasingly complex to create and interpret. But it can be done :) 

Interpreting the coefficients of interaction terms directly from the regression output can be non-intuitive because the coefficients of the main effects and the interaction term need to be considered together. This is where functions like `predict` and packages like `marginaleffects` or `margins` become extremely useful. We touched upon predict in the past but will cover more of it and marginal effects in a future lab. 

## Dichotomous Variable X Multinominal Variable (more than two categories)

We can create and interpret an interaction model with self-employment status (dichotomous) and political party affiliation (multinominal).
```{r dum-cat}
dum_cat_base <- lm(income ~ self_employed + partyid,
                   data = gss)
dum_cat_int <- lm(income ~ self_employed * partyid,
                  data = gss)

modelsummary(list(dum_cat_base, dum_cat_int),
             estimate = "{estimate}{stars}",
             fmt = 3,
             statistic = NULL,
             gof_omit = "IC|Log|alg|pss|F|RMSE",
             notes = "Notes: * p < 0.05, ** p < 0.01, *** p < 0.001",
             output="huxtable")
```


## Dummy X Continuous

We can also create interactions between dummy variables and continuous variables. In the example below, we can hypothesize that income is affected by the hours a person works and whether they have a college degree or higher.

```{r dum-con}
gss <- gss %>% 
  mutate(college = case_when(
    degree %in% c("Bach", "Grad") ~ 1,  # College or Grad
    degree %in% c("Less_HS", "HS", "Assoc") ~ 0   # Non-college
  ))
# table(gss$college, gss$degree)

dum_con_base <- lm(income ~ workhrs + college,
                   data = gss)
dum_con_int <- lm(income ~ workhrs * college,
                  data = gss)

modelsummary(list(dum_con_base, dum_con_int),
             estimate = "{estimate}",
             fmt = 3,
             statistic = NULL,
             gof_omit = "IC|Log|alg|pss|F|RMSE",
             notes = "Notes: * p < 0.05, ** p < 0.01, *** p < 0.001",
             output="huxtable")
```


## Continuous X Continuous

Rounding out the simple cases of interactions, we have continuous X continuous interactions.
```{r con-con}
con_con_base <- lm(income ~ workhrs + educ,
                   data = gss)
con_con_int <- lm(income ~ workhrs * educ,
                  data = gss)

modelsummary(list(con_con_base, con_con_int),
             estimate = "{estimate}",
             fmt = 3,
             statistic = NULL,
             gof_omit = "IC|Log|alg|pss|F|RMSE",
             notes = "Notes: * p < 0.05, ** p < 0.01, *** p < 0.001",
             output="huxtable")
```

## Interactions Between Multiple Variables
You aren't limited to only interacting two variables. You could have 3-way interactions, 4-way interactions, etc. If you find that you need to do this, don't hesitate to reach out. 

