---
title: "Week2_Linear Regression"
author: "Brent H."
date: "2025-01-27"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # FALSE hides the code but shows the output (if any)
knitr::opts_chunk$set(message = FALSE) # FALSE suppresses messages in the output document.
knitr::opts_chunk$set(warning = FALSE) # FALSE suppresses warnings in the output document.

options(scipen=999)  # Set to display small numbers in decimal form
set.seed(1234) # Optional, for reproducibility

# Load required libraries
library(haven) # For reading .dta files
library(tidyverse) # For data manipulation
library(modelsummary) # For really nice descriptive statistics
library(stargazer) # For nice looking tables
library(ggplot2) # For visualizations

```


```{r}
#*******************************
# Loading the Dataset
#*******************************
crime <- read_dta("DATASETS/DC_2000_crime_data.dta")
```

# Exploratory Data Analysis

Viewing the first few rows of the dataset
```{r}
view(crime)
```


Question: Provide a general characterization of the dataset. What kind of information does it contain?
Answer: The dataset contains various socio-economic and crime-related variables for different regions.

Question: Did you notice the small text under the variable names in the data viewer? What does the label say for the variable 'per_nohs'?
Answer: percent of track without a highschool degree

Extracting and learning more about the available variables
Extract variable names
```{r}
var_names <- names(crime)
```


Initialize empty vectors the size of the .... to store variable names and labels
Exercise: Debug the code below:
```{r}
var_labels <- vector("character", length(var_names))

```

Loop through each variable to extract its label
```{r}
for(var in var_names) {
  var_labels[which(var_names == var)] <- attr(crime[[var]], "label")
}
```


Question: Identify the code in the for loop that was used to output the label associated with a variable in the dataset?

Create a data frame with variable names and labels
```{r}
variables_df <- data.frame(
  Variable = var_names,
  Label = var_labels,
  stringsAsFactors = FALSE  # to keep the text as characters
)
```


View the defintion of the available variables in the dataset
```{r}
view(variables_df)

```

Generating summary statistics of the entire data set using 'modelsummary' package by filling in the __BLANK__
```{r}
datasummary_skim(crime)
```


Question: Identify the variable in the dataset that has approximately 4% missing values.
[Students can use the output of datasummary_skim(crime) to answer this question]

Create a histogram for visualizing the distribution of violent crime rate in 2001 by filling in the __BLANK__ with the variable name of interest
```{r}
ggplot(crime, aes(x = vcr2001)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Violent Crime Rate in 2001",
       x = "Violent Crime Rate (per 1,000)",
       y = "Frequency")
```


Exercise: Create a histogram of the violent crime rate in 2001 with bin increments of 25
```{r}
ggplot(crime, aes(x = vcr2001)) +
  geom_histogram(binwidth = 25, fill = "blue", color = "black") +
  labs(title = "Histogram of Violent Crime Rate in 2001",
       x = "Violent Crime Rate (per 1,000)",
       y = "Frequency")
```


Question: How does changing the binwidth to 25 affect the visualization? Is the plot improved?
Answer: Worse, the bars loose information value by being too wide.

Create a density plot of violent crime rate for 2001 by filling in the __BLANK__ with the variable name of interest
```{r}
ggplot(crime, aes(x = vcr2001)) +
  geom_density() +
  geom_vline(aes(xintercept = mean(vcr2001, na.rm = TRUE)), colour="red") +
  labs(title = "Density Plot of Violent Crime Rate in 2001",
       y = "Density",
       x = "Violent Crime Rate (per 1,000)",
       caption = "Notes: Washington D.C. census tracts, 2000") +
  theme(axis.text=element_text(size=12), axis.title=element_text(size=14))
```


Question: What is the numeric value of the red line in the density plot? 
```{r}
mean(crime$vcr2001, na.rm=TRUE)
```
The numeric value of the red line can be found using the mean of 'vcr2001'

# Identify a Bivariate Relationship

Exploring the relationship between % Poverty and Violent Crime Rate
Scatter Plot of Violent Crime Rate and % Poverty
```{r}
crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  ggtitle("Scatter Plot of Violent Crime Rate vs. % Poverty") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty", 
       caption="Notes: Washington D.C. census tracts, 2000") +
  theme_light()
```


Question: What is your impression of the relationship between % poverty and violent crime rate? Is it positive, negative, or non-linear?
Answer: weak positive

Visual Representation of Total Sum of Squares (TSS)
```{r}
mean_vcr <- mean(crime$vcr2001, na.rm = TRUE)

crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  # Add horizontal line representing the overall mean
  geom_hline(yintercept = mean_vcr, linetype = "dashed", color = "blue") +
  # Add vertical lines representing distance from each point to the mean
  geom_segment(aes(xend = poverty_rate, yend = mean_vcr), 
               color = "red", alpha = 0.5) +
  ggtitle("Scatter plot of violent crime rate and % poverty") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty", 
       caption = "Notes: Washington D.C. census tracts, 2000. Red lines represent distances contributing to TSS") +
  theme_minimal()
```

Question: In the above plot, the red lines represent the distance of each observation from the overall mean of violent crime rate. To calculate the Total Sum of Squares (TSS) an important concept for regression analysis, what mathematical operation should you perform on these distances?  What does TSS tell us about the data's variability?

Answer: Square the lengths and sum to get the total variability of the data from its mean. TSS quantifies the total variability in data by measuring the spread of all observations around the mean. A higher TSS indicates greater overall variability available to be explained.

Exercise: Complete the scatter plot with line of best fit and averages for % poverty and violent crime rate. Fill in the __BLANK__ with appropriate code
```{r}
crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001))+
  geom_point(color="black") +
  geom_smooth(method="lm", color="red", fill=NA)+
  geom_vline(data=crime, aes(xintercept = mean(poverty_rate)), colour="blue", lty="dashed") +
  geom_hline(data=crime, aes(yintercept = mean(vcr2001)), colour="blue", lty="dashed") +
  ggtitle("Scatter plot of violent crime rate and % poverty") +
  labs(y="VCR (per 1,000)", x="% Poverty", caption="Notes: Washington D.C. census tracts, 2000 ") 
```


# Exploring the relationship between % Homeownership and Violent Crime Rate

Exercise: Create a Scatter Plot of Violent Crime Rate and % Homeownership with line of best fit and averages
```{r}
crime %>%
  ggplot(aes(x = homeownership_rate, y = vcr2001))+
  geom_point(color="black") +
  geom_smooth(method="lm", color="red", fill=NA)+
  geom_vline(data=crime, aes(xintercept = mean(homeownership_rate, na.rm = TRUE)), colour="blue", lty="dashed") +
  geom_hline(data=crime, aes(yintercept = mean(vcr2001, na.rm = TRUE)), colour="blue", lty="dashed") +
  ggtitle("Scatter Plot of Violent Crime Rate vs. % Homeownership with Averages") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Homeownership", caption="Notes: Washington D.C. census tracts, 2000 ") 
```


Question: After creating the plot, describe the relationship between % homeownership and violent crime rate. 
Answer: Weak negative 


# Simple Linear Regression

Fitting a linear model: Using the poverty_rate to predict the violent crime rate
```{r}
lm_model <- lm(vcr2001 ~ poverty_rate, data = crime)

```

Displaying the summary of the linear model
Exercise: Debug the code below
```{r}
summary(lm_model)

```

Question: Based on the above output, what is the estimated change in violent crime rate for each additional unit increase in poverty rate (poverty_rate)?

Answer: For a one-percentage point increase in poverty rate, we can expect a 0.338 increase in violent crimes per 1,000. In other words, a 10% increase in poverty rate, we should expect to see 3.38 more violent crimes per 1,000.

Question: What is the significance of the intercept in this model? What does it represent in the context of violent crime rate? 
Answer: The p-value is less than 0.001, therefore we reject the null hypothesis that there is no relationship between the poverty rate and violent crime rate. 

Exercise: Create another linear model exploring the relationship between % homeownership and violent crime rate
```{r}
lm_model2 <- lm(vcr2001 ~ homeownership_rate, data = crime)
summary(lm_model2)
```


Using stargazer for a more visually appealing presentation of models (especially in reports or presentations)
```{r}
stargazer(lm_model2, type = "text")

```

Question: Report your findings



# Predicted Values and Residuals 

We will use the linear model lm_model (vcr2001 ~ poverty_rate)

Generating Predicted Values
```{r}
crime$predicted_vcr <- predict(lm_model)

```

Histograms of Actual and Predicted Violent Crime Rates
Objective: Create two overlaid histograms to compare the distribution of actual and predicted violent crime rates.
```{r}
ggplot() +
  geom_histogram(data = crime, aes(x = vcr2001, y = ..density..), fill = "blue", alpha = 0.5, binwidth = 1) +
  geom_histogram(data = crime, aes(x = predicted_vcr, y = ..density..), fill = "red", alpha = 0.5, binwidth = 1) +
  labs(title = "Overlay of Actual vs Predicted Violent Crime Rates",
       x = "Violent Crime Rate",
       y = "Density",
       caption = "Note: Blue=Actual Rates, Red=Predicted Rates") +
  theme_minimal()
```


Question: How do the histograms of the actual and predicted violent crime rates compare? What do you think of the model's predictive accuracy?

Plotting Predicted vs Actual Violent Crime Rates
```{r}
ggplot(crime, aes(x = vcr2001, y = predicted_vcr)) +
  geom_jitter(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "blue") + # Add reference line
  labs(title = "Plot of Predicted vs Actual Violent Crime Rates",
       x = "Actual Violent Crime Rates",
       y = "Predicted Violent Crime Rates") +
  theme_minimal()
```


Question: If our model accounted for all the variation in violent crime rates, how would the plot of predicted vs actual violent crime rates look different?

Answer: All the dots would be on the blue dotted line.

Visualizing the Predicted versus Actual for each observation
```{r}
crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  geom_smooth(method="lm", color="red", fill=NA) +
  geom_segment(aes(xend = poverty_rate, yend = predict(lm_model)), color="blue", linetype="dotted", size=0.7) +
  ggtitle("Relationship Between % Poverty and Violent Crime Rate") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty", caption="Notes: Washington D.C. census tracts, 2000") +
  theme_minimal()
```


Question: What do the vertical blue dotted lines represent?
Answer: The blue lines represent the residuals for each observation in the data.

Generating Residuals
```{r}
crime$residual_vcr <- residuals(lm_model)

```

Density Plot with Normal Distribution Overlay
```{r}
ggplot(crime, aes(x = residual_vcr)) +
  geom_density(aes(y = ..density..), color = "red") +
  stat_function(fun = dnorm, args = list(mean = mean(crime$residual_vcr, na.rm = TRUE), 
                                         sd = sd(crime$residual_vcr, na.rm = TRUE)), color = "blue") +
  labs(title = "Density Plot of Residuals with Normal Distribution Overlay (blue)",
       x = "Residuals",
       y = "Density") +
  theme_minimal()
```


Question: How does the distribution of residuals compare to a normal distribution? What implications does this have for our model?

Answer: Significant deviations from the normal distribution, such as skewness or the presence of outliers, may indicate issues with the model. It could suggest that the model is not capturing all the patterns in the data, there might be unaccounted-for variables influencing the response, or that a linear model might not be the best fit for this data.

Plotting Residuals vs Predicted Values with Reference Line at 0 (predicted = actual)
```{r}
ggplot(crime, aes(x = predicted_vcr, y = residual_vcr)) +
  geom_jitter(color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") + # reference line 
  labs(title = "Residuals vs Predicted Values",
       x = "Predicted Violent Crime Rates", y = "Residuals") +
  theme_minimal()
```


Question: Analyze the Residual vs Predicted plot. Are there any patterns or is the distribution random? What does this imply about your model?

Answer: A random scatter implies that the model is doing well in accounting for the variance in the data. Systematic patterns might indicate that the model is missing some explanatory variables or that the relationship between the variables is not purely linear.


# Supplemental Code on Centering

Centering a Variable: The idea behind centering a variable (like poverty rate) is to adjust the data so that its mean becomes zero. This can help in interpreting the intercept of the regression model as it becomes the expected value of the dependent variable when the independent variable is at its average value.

Recoding to center the poverty rate variable
```{r}
crime_centered <- crime %>%
  mutate(centered_poverty_rate = poverty_rate - mean(poverty_rate, na.rm = TRUE))
```

Centering the poverty rate variable shifts the scale so that the mean is now at zero. You can run all your same analyses. The same rules apply, but now your interpretation of the intercept is different...check it out.

```{r}
if (!require("gridExtra")) {
  install.packages("gridExtra")
  library("gridExtra")
}
```


Fit linear models
```{r}
lm_original <- lm(vcr2001 ~ poverty_rate, data = crime)
lm_centered <- lm(vcr2001 ~ centered_poverty_rate, data = crime_centered)
```


Extract coefficients
```{r}
coef_original <- coef(lm_original)
coef_centered <- coef(lm_centered)
```


Create equation text
```{r}
equation_text_original <- paste("y =", round(coef_original[1], 2), "+", round(coef_original[2], 2), "x")
equation_text_centered <- paste("y =", round(coef_centered[1], 2), "+", round(coef_centered[2], 2), "x")
```


Scatter Plot of Violent Crime Rate vs. Original % Poverty
```{r}
plot1 <- crime %>%
  ggplot(aes(x = poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  geom_smooth(method="lm", color="red", fill=NA) +
  annotate("text", x = 20, y = 60, label = equation_text_original, color = "red", size = 4) +
  annotate("segment", x = 0, xend = 0, y = coef_original[1], yend = 50, arrow = arrow(), color = "red") +
  ggtitle("Original % Poverty") +
  labs(y="Violent Crime Rate (per 1,000)", x="% Poverty") +
  theme_light()
```


Scatter Plot of Violent Crime Rate vs. Centered % Poverty
```{r}
plot2 <- crime_centered %>%
  ggplot(aes(x = centered_poverty_rate, y = vcr2001)) +
  geom_point(color="black") +
  geom_smooth(method="lm", color="blue", fill=NA) +
  annotate("text", x = 10, y = 60, label = equation_text_centered, color = "blue", size = 4) +
  annotate("segment", x = 0, xend = 0, y = coef_centered[1], yend = 50, arrow = arrow(), color = "blue") +
  ggtitle("Centered % Poverty") +
  labs(y="Violent Crime Rate (per 1,000)", x="Centered % Poverty") +
  theme_light()
```


Arrange the plots side by side
```{r}
grid.arrange(plot1, plot2, ncol = 2)
```


Explanation: The first plot uses the original poverty rate, where the intercept of the red line indicates the expected violent crime rate when the poverty rate is zero. The second plot uses the centered poverty rate, where the intercept of the blue line now represents the expected violent crime rate when the poverty rate is at its average. 


# Way to Go! End of Week 2 Lab
