---
title: "Week 10 - Count Outcomes"
author: "Brent H."
date: "2025-04-02"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
options(scipen = 10)

# Libraries
library(tidyverse)       # Data manipulation and ggplot2
library(marginaleffects) # Predictions and slopes for GLMs
library(modelsummary)    # Model summaries
library(MASS)            # glm.nb() for Negative Binomial
library(pscl)            # zeroinfl() for Zero-Inflated Models
library(cowplot)         # Arranging multiple plots
library(haven)           # importing .dta files

library(rstanarm)

if(!require(brms)) install.packages("brms")
library(brms)


set.seed(1234) # Optional, for reproducibility
seed <- 1234
```

# 1. Introduction

Count variables---like the number of siblings or the number of children---often violate classical linear model assumptions due to their discrete, non-negative nature and frequent right-skew. This lab explores Poisson, negative binomial, and zero-inflated approaches to model these outcomes properly.

# 2. Data Preparation

We again rely on the GSS (again). Our key count variables are:

-   Number of sexual partners in the last 12 months (`partners`)

-   Number of siblings (`sibs`)

-   Number of children (`childs`)

```{r}
# Write the code or change your working directory to properly load the GSS2022.dta file and assign it to the object gss2022
gss2022 <- read_dta("../DATASETS/GSS/GSS2022.dta")

gss <- gss2022 %>%
  transmute(
    # AGE: Respondent's age in years
    age = ifelse(age < 0, NA_real_, as.numeric(age)),

    # EDUC: Years of education
    educ = ifelse(educ < 0, NA_real_, as.numeric(educ)),

    # SEX: Biological sex recorded/observed (1 = Male, 2 = Female)
    #      Convert to factor with meaningful labels
    sex = factor(
      case_when(
        sex == 1 ~ "Male",
        sex == 2 ~ "Female",
        TRUE     ~ NA_character_
      ),
      levels = c("Male", "Female")
    ),

    # MARITAL: 1=Married, 2=Widowed, 3=Divorced, 4=Separated, 5=Never Married
    marital = factor(
      ifelse(marital %in% 1:5, marital, NA_real_),
      levels = 1:5,
      labels = c("Married", "Widowed", "Divorced", "Separated", "Never Married")
    ),

    # RACE: 1=White, 2=Black, 3=Other
    race = factor(
      case_when(
        race == 1 ~ 1,
        race == 2 ~ 2,
        race == 3 ~ 3,
        TRUE      ~ NA_real_
      ),
      levels = 1:3,
      labels = c("White", "Black", "Other")
    ),

    # PARTNERS: Number of sexual partners in last 12 months
    partners = ifelse(partners < 0, NA_real_, as.numeric(partners)),

    # SIBS: Total number of siblings
    sibs = ifelse(sibs < 0, NA_real_, as.numeric(sibs)),

    # CHILDS: Total number of children ever had
    childs = ifelse(childs < 0, NA_real_, as.numeric(childs)),

    # WTSSNRPS: The post-stratification weight for 2022 GSS
    weight = ifelse(wtssnrps < 0, NA_real_, wtssnrps)
  ) %>%
  # Dropping all rows that are missing any of these key variables (just for our convenience)
  drop_na()

```

## 2.1 Descriptive Statistics

As always, let's look at our descriptive statistics.

```{r}
datasummary_skim(gss,
                 type = "numeric",
                 fmt = 2, # Show 2 decimal places 
                 histogram = T,
                 title = "Sample Descriptive Statistics: Continuous Variables",
                 notes = "Data: 2022 General Social Survey")

datasummary_skim(gss, 
                 type = "categorical",
                 title = "Sample Descriptive Statistics: Categorical Variables",
                 notes = "Data: 2022 General Social Survey",
                 output = "huxtable")

```

***Reflect:*** Do the means and standard deviations of our count
variables suggest potential skew?

How prevalent are zeros in partners, sibs, or childs?

## 2.2 Exploratory Plots

Visualize the distribution of these count variables to see
zero-inflation or heavy right tails:

```{r}
ggplot(gss, aes(x = partners)) +
  geom_histogram(binwidth = 1, color = "white") +
  theme_minimal() +
  labs(title = "Sexual Partners", x = "partners")

ggplot(gss, aes(x = sibs)) +
  geom_histogram(binwidth = 1, color = "white") +
  theme_minimal() +
  labs(title = "Siblings", x = "siblings")

ggplot(gss, aes(x = childs)) +
  geom_histogram(binwidth = 1, color = "white") +
  theme_minimal() +
  labs(title = "Children", x = "children")


```

***Reflect:***

Which of these looks the most heavily skewed?

How many zeros appear?

# 3. Poisson Regression: Number of Sexual Partners

Poisson regression assumes the mean and variance of the outcome are
equal (i.e., no overdispersion).

##3.1 Model Specification

Below, we predict partners as a function of demographic factors:

```{r}

poi_mod <- glm(
  partners ~ marital + age + educ + sex,
  data = gss,
  weights = weight,  # if your dataset uses survey weights
  family = "poisson"
)

modelsummary(
  list(
    "Log of expected count" = poi_mod,
    "IRRs"      = poi_mod
  ),
  estimate = "{estimate}{stars} ({std.error})",
  exponentiate = c(FALSE, TRUE),
  statistic = NULL,
  gof_omit = "AIC|BIC|Log.Lik|Dev",
  output = "huxtable"
)

```

An IRR over 1.0 suggests a variable increases the rate; an IRR below 1.0 suggests it decreases the rate, compared to the reference.

## 3.2 Check Model Fit: Look at the deviance and degrees of freedom:

```{r}
with(poi_mod, cbind(
  Residual_Deviance = deviance,
  DF_Residual       = df.residual,
  Ratio             = deviance / df.residual
))

```

A ratio of deviance to degrees of freedom near 1 indicates a good Poisson fit. Ratios well above 1 suggest overdispersion.

## 3.3 Predictions and Visualizations

```{r}
plot_predictions(poi_mod, condition = c("age", "sex")) +
  labs(
    title = "Predicted Number of Sexual Partners",
    y = "Predicted Partners"
  )

```

Or see how predictions vary by marital status:

```{r}
plot_predictions(poi_mod, condition = c("marital")) +
  labs(title = "Partners ~ Marital Status & Education")
```

**Exercise:** Calculate the predicted number of sexual partners for a 30-year-old married male with 16 years of education using `predictions`.

```{r}
predictions(poi_mod, newdata = data.frame(marital = "Married", age = 30, 
                                     educ = 16, sex = "Male"), 
       type = "response")
```


#4. Negative Binomial Regression: Number of Siblings When the Poisson

When variance exceeds the mean in count data (overdispersion), Negative Binomial offers an alternative that includes a dispersion parameter

## 4.1 Model Specification

```{r}
nb_mod <- MASS::glm.nb(
  sibs ~ marital + age + educ + sex,
  data = gss,
  weights = weight
)

modelsummary(
  list(
    "Log of expected count" = nb_mod,
    "IRRs" = nb_mod
  ),
  estimate = "{estimate}{stars} ({std.error})",
  exponentiate = c(FALSE, TRUE),
  statistic = NULL,
  gof_omit = "AIC|BIC|Log.Lik|Dev",
  output = "huxtable"
)

```

## 4.2 Overdispersion Check

Run:
```{r}
with(nb_mod, cbind(
  NegBin_Deviance = deviance,
  DF_Residual     = df.residual,
  Ratio           = deviance / df.residual
))
```

If the ratio is near 1, the Negative Binomial model typically is a
better fit than the Poisson in the presence of overdispersion.

## 4.3 Predictions and Visualizations

```{r}
plot_predictions(nb_mod, condition = c("age", "sex")) +
  labs(
    title = "Negative Binomial: Predicted # of Siblings",
    y     = "Predicted Number of Siblings"
  )
```

**Exercise:**
How many siblings is a male with a high school education at ages 20, 30, 40, and 50 expected to have?

```{r}

predictions(nb_mod, newdata = datagrid(age = c(20,30,40,50), educ = 12, sex = "Male"))

```

#5. Zero-Inflated Models: Number of Children

Zero-inflated models are helpful if there are "extra" zeros---some respondents truly never have the event for a "structural" reason, beyond what a single Poisson/NB process would predict.

## 5.1 Model Specification

```{r}
zip_mod <- pscl::zeroinfl(
  childs ~ marital + age + educ + sex | marital + age + educ,
  data   = gss,
  weights= weight,
  dist   = "poisson",   # or "negbin"
  link   = "logit"
)

modelsummary(
  list(
    "log of expected count" = zip_mod,
    "IRR/Odds Ratio" = zip_mod
  ),
  estimate = "{estimate}{stars} ({std.error})",
  exponentiate = c(FALSE, TRUE),
  statistic = NULL,
  gof_omit = "AIC|BIC|Log.Lik|Dev",
  output = "huxtable"
)


```

Here, the formula has two parts:

*Count process:* `childs ~ marital + age + educ + sex`

*Zero-inflation process:* `| marital + age + educ`

***Reflect:***

In the count portion, interpret the coefficient on `age`.

In the zero portion, interpret `educ`. Is more education associated with a higher or lower probability of being a structural zero?

## 5.2 Model Comparison

You can compare ZIP vs. Zero-Inflated Negative Binomial (ZINB):

```{r}
zinb_mod <- pscl::zeroinfl(
  childs ~ marital + age + educ + sex | marital + age + educ,
  data = gss,
  dist = "negbin"
)

# Compare likelihood
lmtest::lrtest(zip_mod, zinb_mod)

```

## 5.3 Visualizing Zero-Inflated Predictions

```{r}
plot_predictions(zip_mod, condition = c("age", "marital")) +
  labs(title = "Zero-Inflated Poisson: Predicted # of Children")

```

***Discuss:***

Does higher education (in the zero model) reduce the likelihood of having children at all?

Does it also reduce the predicted number among those who do have children?

# 6. Putting It All Together

-   Poisson fits count data with moderate means and variances.
-   Negative Binomial is used if you have overdispersion.
-   Zero-Inflated models if there are "excess zeros."

# 7. Bayesian Models

For Bayesian modeling, as usual, the `stan_glm()` command works out similar to the usual `glm()`. For Poisson models, the `family = "poisson"` argument is also the same. Two quick differences:

1.  The big difference is that negative binomial models take the form
    `family = "neg_binomial_2"` instead of using `glm.nb()` as we did
    earlier.
2.  Zero-inflated Bayesian models are not currently supported in
    `rstanarm` but can be implemented in `brms`. However, it poses some
    complications. As such, we'll use the count model (Poisson) to
    demonstrate.

If you want to know more, you can follow this link to Aki Vehtari's page, where he shows how to do this modeling:
<https://avehtari.github.io/modelselection/roaches.html>

```{r bayesmods}
partner_stan_poisson <- stan_glm(partners ~ marital + educ + sex + age + I(age^2),
                                 data = gss, family = "poisson",
                                 seed = seed, chains = 1, refresh = 0)

modelsummary(list("Partners" = partner_stan_poisson),
  estimate = "{estimate}{stars} ({std.error})",
  exponentiate = TRUE,
  statistic = NULL,
  gof_omit = "AIC|BIC|Log.Lik|Dev",
  output = "huxtable")

sibs_stan_negbi <-      stan_glm(sibs ~ marital + educ + sex + age,
                                 data = gss, family = "neg_binomial_2",
                                 seed = seed, chains = 1, refresh = 0)

summary(sibs_stan_negbi)


child_stan_poisson <-   stan_glm(childs ~ marital + educ + sex + age,
                                 data = gss, family = "poisson",
                                 seed = seed, chains = 1, refresh = 0)

summary(child_stan_poisson)

```

Since count data is subject to often-unusual distributions, let's plot our posterior distributions and compare them to the original data.

```{r bayes-plots}
pp_check(partner_stan_poisson) + ggtitle("Partners, Poisson") + scale_x_continuous(breaks = seq(0,10))
pp_check(sibs_stan_negbi) + ggtitle("Siblings, Neg Binom")
pp_check(child_stan_poisson) + ggtitle("Children, Poisson") + scale_x_continuous(breaks = seq(0,10))
```

As we can see, our model performed exceptionally well on predicting siblings with the negative binomial distribution. For partners, it tended to predict more heavily on 1 than the actual distribution. For children, it actually did fairly well, except for missing the mark on those with only one child. I expect that this is due to the zero-inflated nature of the data, which is not accounted for well in the Poisson model.
